\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-05-10 forwards-and-backwards.tex
peter edits done

Most models do not distinguish between juveniles and adults, so,
for example, the number of adults produced by a single parent is determined
by the degree of crowding at the the location of the parent. The novelty
of the models that we introduce here is that, although we shall only
follow the adult population, in formulating the dynamics of the
models we shall distinguish
between production of juveniles, which will depend upon the location of 
the adult, and their successful establishment, which will depend on the
location in which a juvenile lands. The result is that not only the absolute 
number, but also the spatial distribution
around their parent, 
of those offspring that survive to adulthood
will depend upon the local population 
density.

@@

In general this is an extremely difficult question. However, in specical
circumstances, some progress can be made. One of the settings that
has received a great deal of attention in recent years is that in which 
a population is expanding into new territory as a travelling wave. 
Typically one considers a stationary wavefront moving across $\IR^1$, and
most work has focussed on the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\Big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. The first is 
appropriate for modelling the range of an expanding population,
the second for the spread of a selectively advantageous 
mutation through an established (selectively neutral) population. 
Because of the logistic control of the growth rate,
individuals in the wavefront can have much greater reproductive success
than those in the `bulk' of the expanding population. 
Tracing backwards in time, individuals sampled
from behind the wave are `caught' by the wave, upon which they become
trapped in the wavefront. If the population density is very high, then over
suitable timescales, the genealogical trees will be dominated by 
periods of extremely rapid coalescence corresponding to a significant 
proportion of the individuals in the front being descended from a particularly
reproductively successful ancestor. For a surprisingly wide range of models, 
it is believed that, for very large $N$ and on suitable timescales, 
genealogies converge to a 
Bolthausen-Sznitman coalescent, e.g.~\cite{brunet/derrida/mueller/munier:2007}. 
On the other hand, if one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
$$wF(w)=w(1-w)(Cw-1),$$
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from `pulled' to `pushed', \cite{birzu/hallatschek/korolev:2017}.
In that setting, the genealogies will also be quite different
from the Fisher-KPP case. For example, \cite{etheridge/penington:2020}
shows that for a discrete space model corresponding to this 
nonlinearity with $C>4$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
\todo[inline]{Is that condition $C>4$ correct? the paper with Sarah is done for the equation $\partial_t u = \Delta u + u(1-u)(2u-1+\alpha)$ and valid for $\alpha \in (0,1)$. Factorising we would get that equation would be $\partial_t u = \Delta u + (1-\alpha) u(1-u)(\frac{2}{1-\alpha}u-1)$. Supposing the $(1-\alpha)$ in front does not affect the result (which I suppose wouldn't) we get that $C= \frac{2}{1-\alpha}$ and so $\alpha \in (0,1)$ if and only if $C>2$ ...}
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet,
it is typically in a location where population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

@@

In the paper we study scaling limits of spatial popultaions,
obtaining convergence of both the population process
(i.e., the population density as a function of time,
although strictly speaking it is a measure that may not have a density)
as well as of lineages traced back through such a population.
We retain linages through the rescaling process
by means of a lookdown construction.
First, in section~\ref{sec: Model and main results},
we describe the model and the main theorems
(Theorems~\ref{thm:nonlocal_convergence}, \ref{thm:local_convergence}, and~\ref{thm:lineages}).
Next, in Section~\ref{sec:applications}, we discuss a few striking consequences of these results,
namely, X Y and Z \comment{TODO}.
In section \ref{sec:heuristics}, we provide heuristic explanations
of why the theorems ought to be true,
and some key ideas behind them.
In Section \ref{sec:lookdown} we define and discuss the lookdown process that retains the levels.
This is followed by the technical proofs in section \ref{sec:proofs}.

@@

@@

\begin{definition}[Model parameterization and assumptions]
\label{def:model_setup}
Suppose $\gamma(x, m)$, $r(x, m)$, and $F(x, m)$
are bounded, uniformly continuous functions
from $\IR^d \times \IR_{\ge 0}$ to $\IR_{\ge 0}$.
Furthermore, suppose $r$ takes values in $[0, 1]$
and has uniformly bounded second derivatives,
and that $(x, m) \mapsto m^2 \gamma(x, m)$ is unformly bounded.
Fix nonnegative values $\epsilon_r$, $\epsilon_\gamma$, and $\epsilon_F$,
with $\epsilon_r < \epsilon_\gamma$.
Suppose $F: \IR^d \times \IR_{\ge 0} \to \IR$ is bounded and uniformly continous
and define
\begin{align} \label{eqn:mu_defn}
    \mu(x, \eta)
    =
    r(x, \smooth{r} \eta(x))
    \gamma(\smooth{\gamma} \eta(x))
    - \frac{1}{\theta}
    F(x, \smooth{F} \eta(x)) .
\end{align}
\end{definition}

Expressions like $\gamma(x, \smooth{\gamma} \eta(x))$ will appear often in the paper.
To make formula more readable, we overload notation to define
$$
    \gamma(x, \eta) := \gamma(x, \smooth{\gamma} \eta(x)) ,
$$
and similiarly write $\mu(x, \eta)$  for $\mu(x, \smooth{\mu} \eta(x))$ and $r(x, \eta)$ for $r(x, \smooth{r} \eta(x))$.
When convenient, we may also suppress the arguments completely,
writing simply $\gamma$, $\mu$, and $r$ for these quantities.

For each $N > 0$ and $\theta \ge 1/(\epsilon_\gamma - \epsilon_r)$,
the process $\left(\eta^\theta(t)\right)_{t \ge 0}$
is the continuous-time measure-valued Markov process described above
with birth rates, establishment probabilities, and death rates
determined as follows.
When the state of the process is $\eta \in \measures$,
offspring of mass $1/N$ are produced at location $x$
with rate $\theta \gamma(x, \eta) N \eta_s(dx)$
and adults die at location $x$ with rate $\theta \mu(x, \eta) N \eta_s(dx)$,
(i.e., at rates $\theta \gamma$ and $\theta \mu$ per capita respectively).
Each offspring disperses to a location offset from the parent's location
by an independent Gaussian with mean $\meanq(x) / \theta$
and covariance matrix $\covq(x) / \theta$;
an offspring at $x$ establishes instantaneously
with probability $r(x, \eta)$, or else dies.

@@

\begin{definition}[Martingale Problem Characterisation]
    \label{Def: Rescaled Models}
We define $(\eta^N_t)_{t \geq 0}$ to be the solution to the Martingale Problem,
with initial condition $\eta^N_0 \in \measures$,
such that for all smooth functions $f \in C^{2}_{0}(\mathbb{R}^d)$,
writing $\langle f, \eta \rangle = \int_{\IR^d} f(x) \eta(dx)$
and $q_\theta(x, dy)$ as the Gaussian kernel
with mean $x + \meanq(x)/\theta$ and covariance $\covq(x) / \theta$,
\begin{equation}
    \label{eq: Pre-limit Martingale Characterisation}
\begin{aligned}
M^N_t(f)
&:=  \langle f, \eta^N_t \rangle
        -\langle f, \eta^N_0 \rangle
 \\ &\qquad {}
 -  \int_{0}^{t}\bigg\{
        \int\left( \int \theta
             \left(
                f(z)r(z, \eta^N_{s})
                - f(x)r(x, \eta^N_{s})
            \right)
        q_\theta(x,dz) \right)
\\ & \qquad \qquad \qquad \qquad {}
        \times \gamma(x, \eta^N_{s}) \eta^N_{s}(dx)
\\ & \qquad \qquad {}
    + \int f(x) F(x, \eta^N_{s}) \eta^N_{s}(dx)
    \bigg\} ds
\end{aligned}    
\end{equation}
is a martingale with quadratic variation
\begin{equation}
\begin{aligned} \relax
[ M^N_t(f)] =& 
\frac{\theta}{N} \int_{0}^{t}\bigg\{
    \langle \gamma(x, \eta^N_{s})
        \int f^2(y)r(y, \eta^N_{s}) q_\theta(x,dy) 
    , \eta^N_{s}(dx) \rangle \\
& \qquad \qquad {}
    + \langle \mu(x, \eta^N_{s}) f^2(x) 
    , \eta^N_{s}(dx)\rangle
    \bigg\}ds. 
\end{aligned}    
\end{equation}
\end{definition}

@@

\begin{theorem} \label{thm:nonlocal_convergence}
    Let $(\eta^\theta_t)_{t \geq 0}$
    be as defined in Definition \ref{def:model_setup}
    and assume that as $N \to \infty$, $N \to \infty$
    in such a way that $\theta/N \to \alpha$.
    (However, interaction distances $\epsilon_r$, $\epsilon_\gamma$, and $\epsilon_F$
    remain constant.)

    The sequence of stochastic processes
    $(\eta^\theta_t)_{t \geq 0}$
    converges along subsequences in distribution as $N \to \infty$
    to a measure-valued process $(\eta_t)_{t \geq 0}$
    which satisfies that, for every $f:\mathbb{R}^d \rightarrow \mathbb{R}$,
    \begin{align} \label{eqn:mgale_problem}
        M_t
        &:=
        \langle f(x), \eta_t(dx) \rangle
        -
        \langle f(x), \eta_0(dx) \rangle
        \\ & \qquad
        -
        \int_0^t \big\langle
            \gamma(x, \smooth{\gamma} \eta_{s}(x))
            \mathcal{B}\left(
                f(\cdot) r(\cdot, \smooth{r} \eta_{s}(\cdot))
            \right)(x)
            +
            f(x)
            F(x, \smooth{F} \eta_{s}(x)),
            \eta_{s}(dx)
        \big\rangle ds,
    \end{align}
    is a martingale.
    The quadratic variation of $M_t$, denoted $\langle M \rangle_t$, is given by 
    \begin{align} \label{eqn:mgale_variation}
        \langle M \rangle_t
        =
        \alpha
        \int_0^t
        \big\langle
            \left(\gamma\left( x,  \eta_{s} \right)
            r\left(x, \eta_{s} \right)+\mu\left(x, \eta_{s} \right)\right)
            f^2(x),
            \eta_{s} (dx)
        \big\rangle ds. 
    \end{align}
    In other words, the limit is a measure-valued process with non-local interactions.
    Furthermore, if $\alpha = 0$ the limit is deterministic.
\end{theorem}

@@

\begin{corollary}[Superprocess limits] \label{cor:superprocess_uniqueness}
    In the context of Theorem~\ref{thm:nonlocal_convergence},
    if the martingale problem
    defined by equations \eqref{eqn:mgale_problem} and \eqref{eqn:mgale_variation}
    has a unique solution,
    then $(\eta^\theta(t))_{t \ge 0}$ converges to that solution
    as $N \to \infty$.
    \comment{
        Is this what we want to say here? Could be said better.
        TODO: see what Dawson \& Li say about uniqueness
        so we have at least some conditions under which this happens.
    }
\end{corollary}

Our results will show that in determinstic limits ``$\eta_t$ solves a PDE'',
but since $\eta_t$ is a measure for which we don't always know there exists a density,
we need to formalize the weak sense in which it solves the PDE.
The notation in the following definition
is written to suggest the case in which $\eta_t(dx)$ has a density $\varphi_t(x)$.

@@

\begin{corollary}[Deterministic limits]
    \label{cor:nonlocal_pde_limits}
    Under the assumptions of Theorem~\ref{thm:nonlocal_convergence},
    if the limit is deterministic (i.e., $\theta / N \to \alpha = 0$),
    then any limit $\eta_t$ is a weak solution to the nonlocal partial differential equation
    \begin{equation} \label{eqn:nonlocal_pde}
        \partial_t \varphi_t(x)
        =
        r\left(x, \smooth{r} \varphi_t(x) \right)
        \DG^* \left[
            \varphi_t(\cdot)
            \gamma\big( x, \smooth{\gamma} \varphi_t(\cdot) \big)
        \right](x)
        +
        \varphi_t(x)
        F\left(x, \smooth{F} \varphi_t() \right)
        ,
    \end{equation}
    in the sense of Definition~\ref{defn:weak_solutions}.
    If equation \eqref{eqn:nonlocal_pde} has a unique solution,
    then $(\eta^\theta_t)_{t \ge 0}$ converges to that solution.
\end{corollary}

@@

\begin{theorem}[Convergence to a PDE]
    \label{thm:local_convergence}
    Let $(\eta^\theta_t)_{t \geq 0}$
    be as defined in Definition \ref{def:model_setup}
    and assume that as $N \to \infty$, $\theta \to \infty$
    and $\epsilon_r + 1/\theta < \epsilon_\gamma = \epsilon_F = \epsilon \to 0$
    in such a way that $\theta/N \to 0$
    and
    \comment{TODO: some conditions on $\epsilon$ and $r$ and $\gamma$ and $F$}.
    Then the sequence of stochastic processes $(\eta^\theta_t)_{t \ge 0}$
    converges \comment{in some sense}
    to a measure-valued process with a density $\varphi(t, x)$
    that solves
    \begin{align}
        \partial_t \varphi(t, x)
        &=
        \text{whatever the final form is here} .
    \end{align}
\end{theorem}

@@

\comment{
    This section:
    model set-up including the limit;
    comparison of the simulation profile to deterministic solution.
    See Figure \ref{fig:pme_waves}.
}

Consider the porous medium equation (PME) with logistic growth,
\begin{equation}
    \label{eqn:pme}
    \partial_t v = \partial_x^2 (v^2) + v (1 - v) ,
\end{equation}
an example of a reaction-diffusion equation with a nonlinear diffusion term.
Such equations are widely used in a number of contexts in biology in which
it is clear that motility within a population varies with population density.
For example, density dependent dispersal is a common feature in spatial
models in ecology, eukaryotic cell biology, and avascular tumour growth;
see e.g.~\cite{sherratt:2010} and references therein for further discussion. 
We shall pay particular attention to the case in which the equation can be 
thought of as modelling the density of an expanding population. 
The specific equation that we use to illustrate the implications of
our results is one of a class used to study pattern formation in 
bacterial colonies. In particular, it has been suggested as a model for
the expansion of bacteria of the type {\em Paenbacillus dendritiformis} on
a thin layer of agar in a petri dish, 
\cite{cohen/golding/kozlovsky/benjacob/ron:1999}.

Comparing \eqref{eqn:pme} to \eqref{eqn:pde},
we see that to set up a limit in which the population density $\varphi$ follows the PME,
we need $r=1$,
$\gamma = \varphi$, and $F = 1 - \varphi$.
Consulting equation~\eqref{eqn:mu_defn},
we then need death rate to be $\mu = (1 + 1/\theta) \varphi - 1/\varphi$.
In other words,
establishment is certain
and birth rates to increase linearly with population density,
but to compensate, death rates increase slightly faster (also linearly),
as shown in Figure~\ref{fig:pme_waves}.
Alert readers will notice that this violates our boundedness assumptions;
\comment{(perhaps this is too much detail here)}
in practice we choose a suitably small value for $\beta$ and define
\begin{align*}
    \gamma(x, m) &= (1 - \exp(- \beta m)) / \beta \\
    F(x, m) &= \exp(- \beta m) / \beta ,
\end{align*}
which in turn implies that death rate when local density is $m$
is roughly $(1 + 1/\theta) (1 - \exp(-\beta m)) / \beta - 1/(\theta \beta)$.
Birth and death rates are equal at density $m = 1$,
corresponding to an unscaled density of $N$ individuals per unit area.

@@

\comment{
    This section:
    explain generator in stationary frame;
    derive stationary distribution;
    introduce FKPP;
    compute for PME and FKPP.
    (Also: Allen-Cahn example?)
    The two cases are very different, as shown in Figure \ref{fig:pme_vs_fkpp}.
}

It is interesting to investigate the motion of an ancestral lineage
in special cases in which we have an explicit expression for the travelling
wave profile $w$.

%%%%%%%%%%
\paragraph{Porous Media:}
Setting $x_0=0$ (for definiteness) and substituting
the form of $w^P$ from equation \eqref{eqn:pme_wave}
into Corollary~\ref{cor:wavefront},
with $c=1$,
$\gamma(x, m) = m$,
$r(x,w) = 1$,
and $F(x, m) = (1 - m)$,
the lineage's generator is, for $x < 0$,
\comment{(need to decide whether to use $f_x$ or $\delta_x f$)}
\begin{align*}
    \Lgen f
    &=
        w(x)
        \left(
         f_{xx}
         +
         2 \frac{(w^2)_x}{w^2} f_x
        \right)
        + f_x \\
    &=
        \left(1 - e^{\frac{1}{2} x} \right)
        f_{xx}
        -
        2 e^{\frac{1}{2} x} f_x
        +
        f_x .
\end{align*}
The speed measure corresponding to this diffusion is,
again for $x < 0$,
\begin{align*}
    m(d\xi)
    &\propto
        \frac{ 1 }{ 2 (1 - \exp{\xi/2}) }
        \exp\left(
            \int_\eta^\xi \left\{
                1 - \frac{e^{x/2}}{1 - e^{x/2}}
            \right\} dx
        \right) \\
    &\propto
        e^\xi\left(1-e^{\xi/2}\right),
        \quad
        \text{for } \xi < 0 ,
\end{align*}
which is integrable and so when suitably normalised gives the unique stationary distribution.
Notice that at stationarity
the lineage will typically be significantly behind the front.

>>>>

Most models do not distinguish between juveniles and adults, so,
for example, the number of adults produced by a single parent is determined
by the degree of crowding at the location of the parent. The novelty
of the models that we introduce here is that, although we shall only
follow the adult population, in formulating the dynamics of the
models we shall distinguish
between production of juveniles, which will depend upon the location of 
the adult, and their successful establishment, which will depend on the
location in which a juvenile lands. The result is that not only the absolute 
number, but also the spatial distribution
around their parent, 
of those offspring that survive to adulthood
will depend upon the local population 
density.

@@

In general this is an extremely difficult question. However, in special
circumstances, some progress can be made. One of the settings that
has received a great deal of attention in recent years is that in which 
a population is expanding into new territory as a travelling wave. 
Typically one considers a stationary wavefront moving across $\IR^1$, and
most work has focussed on the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\Big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. The first is 
appropriate for modelling the range of an expanding population,
the second for the spread of a selectively advantageous 
mutation through an established (selectively neutral) population. 
Because of the logistic control of the growth rate,
individuals in the wavefront can have much greater reproductive success
than those in the `bulk' of the expanding population. 
Tracing backwards in time, individuals sampled
from behind the wave are `caught' by the wave, upon which they become
trapped in the wavefront. If the population density is very high, then over
suitable timescales, the genealogical trees will be dominated by 
periods of extremely rapid coalescence corresponding to a significant 
proportion of the individuals in the front being descended from a particularly
reproductively successful ancestor. For a surprisingly wide range of models, 
it is believed that, for very large $N$ and on suitable timescales, 
genealogies converge to a 
Bolthausen-Sznitman coalescent, e.g.~\cite{brunet/derrida/mueller/munier:2007}. 
On the other hand, if one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
$$wF(w)=w(1-w)(Cw-1),$$
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from `pulled' to `pushed', \cite{birzu/hallatschek/korolev:2017}.
In that setting, the genealogies will also be quite different
from the Fisher-KPP case. For example, \cite{etheridge/penington:2020}
shows that for a discrete space model corresponding to this 
nonlinearity with $C>4$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
\todo[inline]{Is that condition $C>4$ correct? the paper with Sarah is done for the equation $\partial_t u = \Delta u + u(1-u)(2u-1+\alpha)$ and valid for $\alpha \in (0,1)$. Factorising we would get that equation would be $\partial_t u = \Delta u + (1-\alpha) u(1-u)(\frac{2}{1-\alpha}u-1)$. Supposing the $(1-\alpha)$ in front does not affect the result (which I suppose wouldn't) we get that $C= \frac{2}{1-\alpha}$ and so $\alpha \in (0,1)$ if and only if $C>2$ ...}
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet,
it is typically in a location where population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

@@

In the paper we study scaling limits of spatial populations,
obtaining convergence of both the population process
(i.e., the population density as a function of time,
although strictly speaking it is a measure that may not have a density)
as well as of lineages traced back through such a population.
We retain linages through the rescaling process
by means of a lookdown construction.
First, in section~\ref{sec: Model and main results},
we describe the model and the main theorems
(Theorems~\ref{thm:nonlocal_convergence}, \ref{thm:local_convergence}, and~\ref{thm:lineages}).
Next, in Section~\ref{sec:applications}, we discuss a few striking consequences of these results,
namely, X Y and Z \comment{TODO}.
In section \ref{sec:heuristics}, we provide heuristic explanations
of why the theorems ought to be true,
and some key ideas behind them.
In Section \ref{sec:lookdown} we define and discuss the lookdown process that retains the levels.
This is followed by the technical proofs in section \ref{sec:proofs}.

@@

Were not using $\lambda$, but only $N$.

@@

\begin{definition}[Model parameterization and assumptions]
\label{def:model_setup}
Suppose $\gamma(x, m)$, $r(x, m)$, and $F(x, m)$
are bounded, uniformly continuous functions
from $\IR^d \times \IR_{\ge 0}$ to $\IR_{\ge 0}$.
Furthermore, suppose $r$ takes values in $[0, 1]$
and has uniformly bounded second derivatives,
and that $(x, m) \mapsto m^2 \gamma(x, m)$ is uniformly bounded.
Fix nonnegative values $\epsilon_r$, $\epsilon_\gamma$, and $\epsilon_F$,
with $\epsilon_r < \epsilon_\gamma$.
Suppose $F: \IR^d \times \IR_{\ge 0} \to \IR$ is bounded and uniformly continuous
and define
\begin{align} \label{eqn:mu_defn}
    \mu(x, \eta)
    =
    r(x, \smooth{r} \eta(x))
    \gamma(\smooth{\gamma} \eta(x))
    - \frac{1}{\theta}
    F(x, \smooth{F} \eta(x)) .
\end{align}
\end{definition}

Expressions like $\gamma(x, \smooth{\gamma} \eta(x))$ will appear often in the paper.
To make formula more readable, we overload notation to define
$$
    \gamma(x, \eta) := \gamma(x, \smooth{\gamma} \eta(x)) ,
$$
and similarly write $\mu(x, \eta)$  for $\mu(x, \smooth{\mu} \eta(x))$ and $r(x, \eta)$ for $r(x, \smooth{r} \eta(x))$.
When convenient, we may also suppress the arguments completely,
writing simply $\gamma$, $\mu$, and $r$ for these quantities.

For each $N > 0$ and $\theta \ge 1/(\epsilon_\gamma - \epsilon_r)$,
the process $\left(\eta^N(t)\right)_{t \ge 0}$
is the continuous-time measure-valued Markov process described above
with birth rates, establishment probabilities, and death rates
determined as follows.
When the state of the process is $\eta \in \measures$,
offspring of mass $1/N$ are produced at location $x$
with rate $\theta \gamma(x, \eta) N \eta_s(dx)$
and adults die at location $x$ with rate $\theta \mu(x, \eta) N \eta_s(dx)$,
(i.e., at rates $\theta \gamma$ and $\theta \mu$ per capita respectively).
Each offspring disperses to a location offset from the parent's location
by an independent Gaussian with mean $\meanq(x) / \theta$
and covariance matrix $\covq(x) / \theta$;
an offspring at $x$ establishes instantaneously
with probability $r(x, \eta)$, or else dies.

@@

\begin{definition}[Martingale Problem Characterisation]
    \label{defn:mgale_construction},
We define $(\eta^N_t)_{t \geq 0}$ to be the solution to the Martingale Problem,
with initial condition $\eta^N_0 \in \measures$,
such that for all smooth functions $f \in C^{2}_{0}(\mathbb{R}^d)$,
writing $\langle f, \eta \rangle = \int_{\IR^d} f(x) \eta(dx)$
and $q_\theta(x, dy)$ as the Gaussian kernel
with mean $x + \meanq(x)/\theta$ and covariance $\covq(x) / \theta$,
\begin{equation}
    \label{eq: Pre-limit Martingale Characterisation}
\begin{aligned}
M^N_t(f)
&:=  \langle f, \eta^N_t \rangle
        -\langle f, \eta^N_0 \rangle
 \\ &\qquad {}
 -  \int_{0}^{t}\bigg\{
        \int\left( \int \theta
             \left(
                f(z)r(z, \eta^N_{s})
                - f(x)r(x, \eta^N_{s})
            \right)
        q_\theta(x,dz) \right)
\\ & \qquad \qquad \qquad \qquad {}
        \times \gamma(x, \eta^N_{s}) \eta^N_{s}(dx)
\\ & \qquad \qquad {}
    + \int f(x) F(x, \eta^N_{s}) \eta^N_{s}(dx)
    \bigg\} ds
\end{aligned}    
\end{equation}
is a martingale with quadratic variation
\begin{equation}
\begin{aligned} \relax
[ M^N_t(f)] =& 
\frac{\theta}{N} \int_{0}^{t}\bigg\{
    \langle \gamma(x, \eta^N_{s})
        \int f^2(y)r(y, \eta^N_{s}) q_\theta(x,dy) 
    , \eta^N_{s}(dx) \rangle \\
& \qquad \qquad {}
    + \langle \mu(x, \eta^N_{s}) f^2(x) 
    , \eta^N_{s}(dx)\rangle
    \bigg\}ds. 
\end{aligned}    
\end{equation}
\end{definition}

@@

\begin{theorem} \label{thm:nonlocal_convergence}
    Let $(\eta^N_t)_{t \geq 0}$
    be as defined in Definition \ref{def:model_setup}
    and assume that as $N \to \infty$, $N \to \infty$
    in such a way that $\theta/N \to \alpha$.
    (However, interaction distances $\epsilon_r$, $\epsilon_\gamma$, and $\epsilon_F$
    remain constant.)

    The sequence of stochastic processes
    $(\eta^N_t)_{t \geq 0}$
    converges along subsequences in distribution as $N \to \infty$
    to a measure-valued process $(\eta_t)_{t \geq 0}$
    which satisfies that, for every $f:\mathbb{R}^d \rightarrow \mathbb{R}$,
    \begin{align} \label{eqn:mgale_problem}
        M_t
        &:=
        \langle f(x), \eta_t(dx) \rangle
        -
        \langle f(x), \eta_0(dx) \rangle
        \\ & \qquad
        -
        \int_0^t \big\langle
            \gamma(x, \smooth{\gamma} \eta_{s}(x))
            \mathcal{B}\left(
                f(\cdot) r(\cdot, \smooth{r} \eta_{s}(\cdot))
            \right)(x)
            +
            f(x)
            F(x, \smooth{F} \eta_{s}(x)),
            \eta_{s}(dx)
        \big\rangle ds,
    \end{align}
    is a martingale.
    The quadratic variation of $M_t$, denoted $\langle M \rangle_t$, is given by 
    \begin{align} \label{eqn:mgale_variation}
        \langle M \rangle_t
        =
        \alpha
        \int_0^t
        \big\langle
            \left(\gamma\left( x,  \eta_{s} \right)
            r\left(x, \eta_{s} \right)+\mu\left(x, \eta_{s} \right)\right)
            f^2(x),
            \eta_{s} (dx)
        \big\rangle ds. 
    \end{align}
    In other words, the limit is a measure-valued process with non-local interactions.
    Furthermore, if $\alpha = 0$ the limit is deterministic.
\end{theorem}

@@

\begin{corollary}[Superprocess limits] \label{cor:superprocess_uniqueness}
    In the context of Theorem~\ref{thm:nonlocal_convergence},
    if the martingale problem
    defined by equations \eqref{eqn:mgale_problem} and \eqref{eqn:mgale_variation}
    has a unique solution,
    then $(\eta^N(t))_{t \ge 0}$ converges to that solution
    as $N \to \infty$.
    \comment{
        Is this what we want to say here? Could be said better.
        TODO: see what Dawson \& Li say about uniqueness
        so we have at least some conditions under which this happens.
    }
\end{corollary}

Our results will show that in deterministic limits ``$\eta_t$ solves a PDE'',
but since $\eta_t$ is a measure for which we don't always know there exists a density,
we need to formalize the weak sense in which it solves the PDE.
The notation in the following definition
is written to suggest the case in which $\eta_t(dx)$ has a density $\varphi_t(x)$.

@@

\begin{corollary}[Deterministic limits]
    \label{cor:nonlocal_pde_limits}
    Under the assumptions of Theorem~\ref{thm:nonlocal_convergence},
    if the limit is deterministic (i.e., $\theta / N \to \alpha = 0$),
    then any limit $\eta_t$ is a weak solution to the nonlocal partial differential equation
    \begin{equation} \label{eqn:nonlocal_pde}
        \partial_t \varphi_t(x)
        =
        r\left(x, \smooth{r} \varphi_t(x) \right)
        \DG^* \left[
            \varphi_t(\cdot)
            \gamma\big( x, \smooth{\gamma} \varphi_t(\cdot) \big)
        \right](x)
        +
        \varphi_t(x)
        F\left(x, \smooth{F} \varphi_t() \right)
        ,
    \end{equation}
    in the sense of Definition~\ref{defn:weak_solutions}.
    If equation \eqref{eqn:nonlocal_pde} has a unique solution,
    then $(\eta^N_t)_{t \ge 0}$ converges to that solution.
\end{corollary}

@@

\begin{theorem}[Convergence to a PDE]
    \label{thm:local_convergence}
    Let $(\eta^N_t)_{t \geq 0}$
    be as defined in Definition \ref{def:model_setup}
    and assume that as $N \to \infty$, $\theta \to \infty$
    and $\epsilon_r + 1/\theta < \epsilon_\gamma = \epsilon_F = \epsilon \to 0$
    in such a way that $\theta/N \to 0$
    and
    \comment{TODO: some conditions on $\epsilon$ and $r$ and $\gamma$ and $F$}.
    Then the sequence of stochastic processes $(\eta^N_t)_{t \ge 0}$
    converges \comment{in some sense}
    to a measure-valued process with a density $\varphi(t, x)$
    that solves
    \begin{align}
        \partial_t \varphi(t, x)
        &=
        \text{whatever the final form is here} .
    \end{align}
\end{theorem}

@@

Consider the porous medium equation (PME) with logistic growth,
\begin{equation}
    \label{eqn:pme}
    \partial_t v = \partial_x^2 (v^2) + v (1 - v) ,
\end{equation}
an example of a reaction-diffusion equation with a nonlinear diffusion term.
Such equations are widely used in a number of contexts in biology in which
it is clear that motility within a population varies with population density.
For example, density dependent dispersal is a common feature in spatial
models in ecology, eukaryotic cell biology, and avascular tumour growth;
see~\cite{sherratt:2010} and references therein for further discussion. 
In particular, it has been suggested as a model for
the expansion of a certain type of bacteria % of the type {\em Paenbacillus dendritiformis}
on a thin layer of agar in a petri dish 
\citep{cohen/golding/kozlovsky/benjacob/ron:1999}. 
We shall pay particular attention to the case in which the equation can be 
thought of as modelling the density of an expanding population.

Comparing \eqref{eqn:pme} to \eqref{eqn:pde},
we see that to set up a limit in which the population density $\varphi$ follows the PME,
we need $r=1$,
$\gamma = \varphi$, and $F = 1 - \varphi$.
Consulting equation~\eqref{eqn:mu_defn},
this implies that $\mu = (1 + 1/\theta) \varphi - 1/\varphi$.
In other words,
establishment is certain
and birth rates increase linearly with population density,
but to compensate, death rates increase slightly faster (also linearly),
as shown in Figure~\ref{fig:pme_waves}.
Alert readers will notice that
for finite $\theta$ death rates may be negative,
and furthermore rates are not bounded;
in practice we choose a suitably small value for $\beta$ and define
\begin{align*}
    \gamma(x, m) &= (1 - \exp(- \beta m)) / \beta \\
    F(x, m) &= \exp(- \beta m) / \beta ,
\end{align*}
and set $\mu(x, m) = \max(0, \gamma - \theta F)$.
Birth and death rates are equal at density $m = 1$,
corresponding to an unscaled density of $N$ individuals per unit area.

@@

It is interesting to investigate the motion of an ancestral lineage
in special cases in which we have an explicit expression for the travelling
wave profile $w$. 
We consider three relatively well-studied examples:
the Porous Media equation (above),
the Fisher-KPP equation,
and the Allen-Cahn equation.
We will work here in one dimension,
and take $\sigma^2 = 1$ and $\meanq = 0$.

%%%%%%%%%%
\paragraph{Porous Media:}
Setting $x_0=0$ (for definiteness) and substituting
the form of $w^P$ from equation \eqref{eqn:pme_wave}
into Corollary~\ref{cor:wavefront},
with $c=1$,
$\gamma(x, m) = m$,
$r(x,w) = 1$,
and $F(x, m) = (1 - m)$,
the lineage's generator is, for $x < 0$,
with $f_x$ denoting the spatial derivative of $f$,
\begin{align*}
    \Lgen f
    &=
        w(x)
        \left(
         f_{xx}(x)
         +
         2 \frac{(w(x)^2)_x}{w(x)^2} f_x(x)
        \right)
        + f_x(x) \\
    &=
        \left(1 - e^{\frac{1}{2} x} \right)
        f_{xx}(x)
        -
        2 e^{\frac{1}{2} x} f_x(x)
        +
        f_x(x) .
\end{align*}
The speed measure corresponding to this diffusion is,
again for $x < 0$,
\begin{align*}
    m(d\xi)
    &\propto
        \frac{ 1 }{ 2 (1 - e^{\xi/2}) }
        \exp\left(
            \int_\eta^\xi \left\{
                1 - \frac{e^{x/2}}{1 - e^{x/2}}
            \right\} dx
        \right) \\
    &\propto
        e^\xi\left(1-e^{\xi/2}\right),
        \quad
        \text{for } \xi < 0 ,
\end{align*}
which is integrable and so when suitably normalised gives the unique stationary distribution.
Notice that at stationarity
the lineage will typically be significantly behind the front.
